\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-07-04 papers/methods/paper.tex
Slight update before I head off. Need to update the plots using the new features in ChainConsumer
Updating from some of Dans feedback

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
\newcommand{\name}{Steve-o}
\newcommand{\myemail}{samuelreay@gmail.com}
\newcommand\abs[1]{\left|#1\right|}
\newcommand {\etal} {\emph{~et~al.} }
\newcommand{\green}{\color{green}}
\newcommand{\blue}{\color{blue}}
\newcommand{\red}{\color{red}}

@@

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account a full range of analysis systematics and selection effects. This advances the work of \citet{Rubin2015} by including additional sources of systematic uncertainty and improved treatment of Malmquist bias whilst increasing numerical efficiency. Our model is optimised for efficient fitting and has been validated on a combined simulated dataset of more than $250\,000$ supernovae. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations of the combined low-redshift and DES 3-year spectroscopic survey sample quantify the bias for flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. SNANA simulations can recover the input cosmology to within $0.27\sigma$, while taking into account many more effects than standard analyses.
\end{abstract}
% Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis.

@@

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed Type Ia supernovae (SNIa) have increased by more than an order of magnitude thanks to modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, Sako2014}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structureand the CMB. For a recent review, see \citet{Huterer2018}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

In attempts to tease out the nature of dark energy, active and planned surveys are once again ramping up their statistical power. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} will be observing thousands of Type Ia supernova, attaining both spectroscopic and photometric selected samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will produce scores of thousands of photometrically classified supernovae. Such increased statistical power demands greater fidelity and flexibility in modelling the supernovae for cosmological purposes, as systematic uncertainty will prove to be the limiting factor in our analyses.

As such, staggering effort is being put into developing more sophisticated supernovae analyses. The role of simulations has become increasingly important in determining biases and validating models. First used in ESSENCE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, simulations are now a fundamental component of modern supernovae cosmology.  \citet{Betoule2014} make use of redshift dependent $\mu$-bias to correct their analyses, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulation corrections as a function of redshift, stretch and colour to better remove fitting biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Hierarchical Bayesian models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, and either also use simulation corrections or attempt to find sufficient analytic approximations for complicated effects such as Malmquist bias.

In this paper, we lay out a new hierarchical model that advances the past work of \citet{Rubin2015} to include more robust treatment of systematics and selection effects whilst improving computational performance. Section \ref{sec:review} is dedicated to a quick review of the supernovae cosmology. In Section \ref{sec:method} we outline our methodology. Model verification on simulated datasets is given in Section \ref{sec:verification}, along with details on potential areas of improvement.

@@

Whilst supernova observations take the form of time-series photometric measurements of brightness in many photometric bands, most analyses do not work from these measurements. Instead, most techniques fit these observations (along with redshift) to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernovae light curves for cosmology \citep{Guy2010, Mosher2014}. The resulting output from the model is, for each supernova, a characterised amplitude $x_0$ (which can be converted into apparent magnitude $m_B = -2.5\log(x_0)$), a stretch term $x_1$ and colour term $c$, along with a covariance matrix describing the uncertainty on these summary statistics. As such, the product at the end is a (redshift-dependent) population of $m_B$, $x_1$ and $c$, where the width of this distribution is called the intrinsic dispersion of the supernova population.

The underlying actual supernova population is not such a clear cut, and accurately characterising this population, its evolution over redshift, and effects from environment, is one of the challenges of supernova cosmology. However, given some modelled underlying population that lives in the redshift-dependent space $M_B$ (absolute magnitude), $x_1$ and $c$, the introduction of cosmology into the model is simple -- it is encoded in the functional map between those two populations, from apparent magnitude space to absolute magnitude. Specifically, for any given supernova our functional map may take the traditional form:

\begin{equation}
M_B = m_B + \alpha x_1 - \beta c - \mu(z) + \text{corrections}, \label{eq:standard}
\end{equation}

where $\alpha$ is the stretch correction \citep{Phillips1993, Phillips1999}, and $\beta$ is the colour correction \citep{Tripp1998} that respectively encapsulate the empirical relation that broader and bluer supernovae are brighter. The $\text{corrections}$ term often includes corrections for host galaxy environment, as this has statistically significant effects on supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013, Uddin2017}. The cosmological term, $\mu(z)$ represents the distance modulus, and is precisely known given cosmological parameters and an input redshift.

@@

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about uncertainty correlation. For example, correlation the underlying supernova colour population's mean and skewness is highly correlated, however this correlation is lost when determining population uncertainty using numerical derivatives of population permutations. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

Selection efficiency, such as the well known Malmquist bias \citep{MalmquistK.G.1922} is accounted for by correcting data. Specifically, Malmquist bias is the result of losing the fainter tail of the supernova population at high redshift when supernovae increasingly fall below the detection threshold. An example of Malmquist bias is illustrated in Figure \ref{fig:malmquist}. Simulations following survey observational strategies and geometry are used to calculate the expected bias in distance modulus, which is then added onto the observational data. As these effects are not built into the likelihood, their influence on the error budget is not captured fully in the $\chi^2$ distribution, and any subtle correlations between cosmological or population parameters and the bias is lost.

@@

By building the systematic uncertainties and selection effects into the simulation package, there is more freedom in how to treat and model those effects. Data does not need to be corrected, analytic approximations do not need to be applied, we are free to incorporate algorithms that simply cannot be expressed in a tractable likelihood. This freedom comes with a cost -- computation. The classical $\chi^2$ method's most computationally expensive step in a fit is matrix inversion. For ABC methods, we must instead simulate an entire supernova population -- drawing from underlying supernova populations, modelling light curves, applying selection effects, fitting light curves and applying data cuts. This is an intensive process.

@@

With the introduction of multiple layers in our model, we can add far more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009, Mandel2011a, Mandel2017} construct a hierarchical model that they apply to supernova light curve fitting. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide increased flexibility with reduced uncertainty over the traditional method, but does not incorporate bias correction. \citet{March2014, Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS like simulations. The recent BAHAMAS model \citep{Shariff2016} builds on this and reanalyses the JLA dataset (using their redshift dependent bias corrections), whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} performed a reanalysis of the JLA dataset within a Bayesian formulation, finding significant differences in $\alpha$ and $\beta$ values from the original analysis from \citet{Betoule2014}. Notably, these methods rely on data that is bias corrected or ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The simplification made by the UNITY analysis is that the bias is well described by an analytic function, and this function is fixed such that selection effects are assumed to be determined with perfect precision. Their model was validated on thirty realisations of statistics.

@@

We construct our hierarchical Bayesian model with several goals in mind: creation of a redshift-dependent underlying supernova population, increased treatment of systematics, and analytic correction of selection effects, including systematic uncertainty on those corrections. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in Stan \citep{Carpenter2017, StanDevelopmentTeam2017}. The primary challenge of fitting hierarchical models is their large number of fit parameters, and Stan, which uses automatic differentiation and the no-U-turn Sampler (NUTS) (a variant of Hamiltonian Monte Carlo), allows us to efficiently sample high dimensional parameter space.

At the most fundamental level, a supernova analysis is simply a mapping from an underlying population onto an observed population, where cosmology is encoded directly in the mapping function. The difficulty arises both in adequately describing the biases in the mapping function, and in adding sufficient, physically motivated flexibility in both of these populations whilst not adding \textit{too} much flexibility, such that model fitting becomes pathological due to increasing parameter degeneracies within the model. In the following sections, we will describe these layers and mapping functions.

@@

For the moment, let us consider a single supernova, and represent its true parameters as $\theta$. The underlying supernova model and intrinsic dispersion model gives the probability of $p(y|\theta)$ for obtaining light curve data $y$. As mentioned earlier, we work with summary statistics -- numbers output from a light curve fitter given the input $y$. The function $f$ describes the action of the fitter $\lbrace \hat{m}_B, \hat{x}_1., \hat{c} \rbrace = f(y)$. For notational simplify, let us denote $\eta \equiv \lbrace m_B, x_1, c \rbrace$. The model prediction for our derived data is

@@

Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift, and subtract this shift from our observed data. With the extra dispersion introduced from the $\kappa$ parameters addressing the under-reporting of error, shifting the mean colour address the non-Gaussianity introduced by chromatic smearing. We allow the amount of shift to be parametrised to account for different intrinsic scatter models, where parameter $s_c$ controls the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations \citep[which corresponds to the intrinsic scatter model from][]{Chotard2011}. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a universal shift in supernovae colour. We place a uniform prior on $s_c$. It is important to note that the effects being modelled and characterised by simulations here are features of the supernova population, not cosmological features, and so the choice of fiducial cosmology has negligible impact.

@@

The underlying supernova population is characterised by underlying distributions in colour, stretch and magnitude. Analytic models often treat the colour and stretch populations independently with a skew normal and normal, respectively. Intrinsic dispersion is treated in a variety of manners in other models, from representing it simply as (Gaussian) scatter on the absolute magnitude of the supernova population, to correlated multivariate normal scatter on the combined magnitude, stretch and colour distribution. Additionally, redshift drift of populations' mean colour and stretch will introduce a cosmological bias in our fits unless the model population possesses similar ability to change as a function of redshift.

We follow prior work and model the colour population as an independent redshift-dependent skew normal for each survey. For the stretch population, we adopt a redshift-dependent normal, and magnitude dispersion is modelled as a normal. Following {\rubin} we allow the mean colour and stretch to vary over redshift, anchoring four equally spaced redshift nodes spanning the redshift range of each survey, linearly interpolating between the nodes. These nodes are represented as $\langle x_1^i \rangle$ and $\langle c^i \rangle$. Both the colour and stretch means are modelled with normal priors. Initial versions of our model adopted a fully covariant multivariate skew normal (with skewness set to zero only for the magnitude component), however pathological fitting complications required us to simplify our treatment. We parameterise the skewness $\alpha_c$ by sampling $\delta_c = \alpha_c / \sqrt{1 + \alpha_c^2}$ which itself is given a uniform prior $\mathcal{U}(0,0.98)$, which allows $\alpha_c$ to span positive values in the realm of physical plausibility as determined from constraints in \citet{Scolnic2016}. The width of the population, represented by the vector $\lbrace \sigma_{M_B}, \sigma_{x_1}, \sigma_c \rbrace$ is subject to Cauchy priors, however are sampled in log space for efficiency in sampling close to zero.

@@

On top of the SNIa populations, as described above, we also include a simplistic outlier population that also follows {\rubin} \citep[and therefore ][]{Kunz2007} as a Gaussian mixture; where the mean of the population is fixed to the SNIa population, but the population width is set to a width of $\sigma^{\rm outl} = 1$ in $M_B$, $x_1$ and $c$. With the spectroscopic DES sample, the contamination rate is expected to be far too low to actually fit contamination population, however in future works with photometric samples that will suffer from significantly more contamination, it will be required that extra degrees of freedom are afforded the outlier population. This contamination is not just a result of non-SNIa being observed, but can also arise from host galaxy misidentification and incorrect redshifts - depending on the cuts on the optimised ratio between purity and efficiency, this can result in between 3\% to 9\% misidentification  \citep{Gupta2016}. Proof of concept analysis on simple simulation data show that an acceptable parametrisation to deal with contamination is to represent the typically dimmer contaminant population as $\langle M_B^{\rm outl} \rangle = \langle M_B \rangle + \delta_{M_B}^{\rm outl}$, where $\delta_{M_B}^{\rm outl}$ is constrained to be positive, or even to be greater than a small positive number to reduce degeneracy between the two populations. For the purposes of the DES spectroscopic sample, which will be confirmed SNIa, $\delta_{M_B}^{\rm outl} =0$. We assume that supernovae fall into the SNIa population as determined by their observed classification probability $\hat{p}$, which is set to unity for the spectroscopic sample.

@@

It is now well known that host galaxy environment has a significant effect on supernova properties. The latest sample of over 1300 spectroscopically confirmed Type Ia supernovae show $>5\sigma$ evidence for correlation between host mass and luminosity \citep{Uddin2017}. The traditional correction, as employed in analyses such as \citet{Suzuki2012} and \citet{Betoule2014} invoke a step function such that $\Delta M = 0.08 \mathcal{H}(\log(M) - 10))$, where $\mathcal{H}$ is the Heaviside step function, $M$ is the galaxy mass in solar masses and $0.08$ represents the size of the magnitude step. The scale of this step function varies from analysis to analysis, with the 0.08 value shown previously sourced from \cite{Sullivan2010} and used in \citet{Betoule2014}. In this work we adopt the model used in {\rubin}, which follows the work from \citet{Rigault2013}, such that we introduce two parameters to incorporate a redshift-dependent host galaxy mass correction:

\begin{equation}
\Delta M = \delta(0) \left[ \frac{1.9\left(1 - \frac{\delta(0)}{\delta(\infty)}\right)  }{0.9 + 10^{0.95z}} + \frac{\delta(0)}{\delta(\infty)}\right], \label{eq:mass}
\end{equation}

@@

This method of adjusting the observed summary statistics is used throughout the traditional and BHM analyses, however it is normally constrained to band systematics. That is, each band for each survey has two systematics associated with it - the calibration uncertainty and the filter wavelength uncertainty. We include these in our approach, in addition to including HST Calspec calibration uncertainty, ten SALT2 model systematic uncertainties, a dust systematic, a global redshift bias systematic and also the systematic peculiar velocity uncertainty. This gives thirteen global systematics shared by all surveys, plus two systematics per band in each survey. With $\eta \equiv \lbrace m_B, x_1, c \rbrace$, our initial conditional likelihood for our observed summary statistics shown in Equation \eqref{eq:pop} becomes

>>>>

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
\newcommand{\name}{Steve}
\newcommand{\myemail}{samuelreay@gmail.com}
\newcommand\abs[1]{\left|#1\right|}
\newcommand {\etal} {\emph{~et~al.} }
\newcommand{\green}{\color{green}}
\newcommand{\blue}{\color{blue}}
\newcommand{\red}{\color{red}}

@@

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account a full range of analysis systematics and selection effects. This advances previous works with hierarchical Bayesian models by including additional sources of systematic uncertainty and improved treatment of Malmquist bias, whilst increasing numerical efficiency. Our model is optimised for efficient fitting and has been validated on a combined simulated dataset of more than $250\,000$ supernovae to rule out the presence of significant model biases in parameter recovery. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations of the combined low-redshift and DES 3-year spectroscopic survey sample quantify the bias for flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. SNANA simulations can recover the input cosmology to within $0.27\sigma$, while taking into an increased number of systematics than prior analyses.
\end{abstract}
% Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis.

@@

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed Type Ia supernovae (SNIa) have increased by more than an order of magnitude, with contributions from modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, Sako2014}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structureand the CMB. For a recent review, see \citet{Huterer2018}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

In attempts to tease out the nature of dark energy, active and planned surveys are once again ramping up their statistical power. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has observed thousands of Type Ia supernova, attaining both spectroscopic and photometric selected samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will observe scores of thousands of photometrically classified supernovae. Such increased statistical power demands greater fidelity and flexibility in modelling the supernovae for cosmological purposes, as systematic uncertainties will prove to be the limiting factor in our analyses \citep{Betoule2014}.

As such, staggering effort is being put into developing more sophisticated supernova cosmology analyses. The role of simulations mimicking survey observations has become increasingly important in determining biases in cosmological constraints and validating specific cosmological models. First used in ESSENCE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, simulations are now a fundamental component of modern supernovae cosmology.  \citet{Betoule2014} quantise and then correct observational bias using simulations, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulation to quantify observational bias in as a function of many factors to better remove fitting biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Hierarchical Bayesian models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, and either also use simulation corrections or attempt to find sufficient analytic approximations for complicated effects such as Malmquist bias.

In this paper, we lay out a new hierarchical model that builds off the past work of \citet{Rubin2015} to include more robust treatment of systematics and selection effects, whilst improving computational performance. Section \ref{sec:review} is dedicated to a quick review of the supernovae cosmology. In Section \ref{sec:method} we outline our methodology. Model verification on simulated datasets is given in Section \ref{sec:verification}, along with details on potential areas of improvement.

@@

Whilst supernova observations take the form of time-series photometric measurements of brightness in many photometric bands, most analyses do not work from these measurements directly. Instead, most techniques fit and observed redshift and these photometric observations to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernovae light curves for cosmology \citep{Guy2010, Mosher2014}. The resulting output from the model is, for each supernova, a characterised amplitude $x_0$ (which can be converted into apparent magnitude $m_B = -2.5\log(x_0)$), a stretch term $x_1$ and colour term $c$, along with a covariance matrix describing the uncertainty on these summary statistics. As all supernova are not identical, an ensemble of supernova form a redshift-dependent, observed population of $m_B$, $x_1$ and $c$, where the width of this distribution is called the intrinsic dispersion of the supernova population.

The population previous discussed represents an observed population, which -- due to the presence of various selection effects -- may not represent the underlying supernova population. Accurately characterising this population, its evolution over redshift, and effects from environment, is one of the challenges of supernova cosmology. However, given some modelled underlying supernova population that lives in the redshift-dependent space $M_B$ (absolute magnitude), $x_1$ and $c$, the introduction of cosmology into the model is simple -- it is encoded in the functional map between those two populations, from apparent magnitude space to absolute magnitude. Specifically, for any given supernova our functional map may take the traditional form:

\begin{equation}
M_B = m_B + \alpha x_1 - \beta c - \mu(z) + \text{other corrections}, \label{eq:standard}
\end{equation}

where $M_B$ is the mean absolute magnitude for all SN Ia  $\alpha$ is the stretch correction \citep{Phillips1993, Phillips1999}, and $\beta$ is the colour correction \citep{Tripp1998} that respectively encapsulate the empirical relation that broader (longer-lasting) and bluer supernovae are brighter. The cosmological term, $\mu(z)$ represents the distance modulus, and is precisely known given cosmological parameters and a redshift. The $\text{other corrections}$ term often includes corrections for host galaxy environment, as this has statistically significant effects on supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013, Uddin2017} In traditional analyses, the other corections term often includes bias corrections, which can take the form of a redshift-dependent function \citep{Betoule2014}.

@@

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about correlation between sources of uncertainty. For example, the underlying supernova colour population's mean and skewness is highly correlated, however this correlation is lost when determining population uncertainty using numerical derivatives of population permutations. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. A common solution to this computational cost problem is to bin the supernova.

Selection efficiency, such as the well known Malmquist bias \citep{MalmquistK.G.1922} is accounted for by correcting data. Specifically, Malmquist bias is the result of losing the fainter tail of the supernova population at high redshift when supernovae increasingly fall below the detection threshold. An example of Malmquist bias is illustrated in Figure \ref{fig:malmquist}. Simulations following survey observational strategies and geometry are used to calculate the expected bias in distance modulus, which is then added onto the observational data. As these effects are not built into the likelihood and instead are formed by correcting data, their uncertainty is not captured fully in the $\chi^2$ distribution, and any subtle correlations between cosmological or population parameters and the bias is lost.

@@

Simulations provide great flexibility and freedom in how to treat the systematic uncertainties and selection effects that plague supernovae surveys. By using forward modelling directly from these simulations, data does not need to be corrected, analytic approximations do not need to be applied; we are free to incorporate algorithms that simply cannot be expressed in a tractable likelihood. This freedom comes with a cost -- computation. The classical $\chi^2$ method's most computationally expensive step in a fit is matrix inversion. For ABC methods, we must instead simulate an entire supernova population -- drawing from underlying supernova populations, modelling light curves, applying selection effects, fitting light curves and applying data cuts. This is an intensive process.

@@

With the introduction of multiple layers in our model, we can add far more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009, Mandel2011a, Mandel2017} construct a hierarchical model that they apply to supernova light curve fitting. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide increased flexibility with reduced uncertainty over the traditional method, but does not incorporate bias correction. \citet{March2014, Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS like simulations. The recent BAHAMAS model \citep{Shariff2016} builds on this and reanalyses the JLA dataset (using their redshift dependent bias corrections), whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} performed a reanalysis of the JLA dataset within a Bayesian formulation, finding significant differences in $\alpha$ and $\beta$ values from the original analysis from \citet{Betoule2014}. Notably, these methods rely on data that is bias corrected or ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The simplification made by the UNITY analysis is that the bias is well described by an analytic function, and this function is fixed such that selection effects are assumed to be determined with perfect precision. Their model was validated to be free of significant biases using fits to thirty realisations of statistics.

@@

We construct our hierarchical Bayesian model with several goals in mind: creation of a redshift-dependent underlying supernova population, increased treatment of systematics, and analytic correction of selection effects, including systematic uncertainty on those corrections. We also desire this to be more computationally efficient than prior works, such that cosmological results from thousands of supernovae are obtainable in the order of hours, rather than days. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in Stan \citep{Carpenter2017, StanDevelopmentTeam2017}. The primary challenge of fitting hierarchical models is their large number of fit parameters, and Stan, which uses automatic differentiation and the no-U-turn Sampler (NUTS) (a variant of Hamiltonian Monte Carlo), allows us to efficiently sample high dimensional parameter space.

At the most fundamental level, a supernova analysis is simply a mapping from an underlying population onto an observed population, where cosmology is encoded directly in the mapping function. The difficulty arises both in adequately describing the biases in the mapping function, and in adding sufficient, physically motivated flexibility in both of these populations whilst not adding \textit{too} much flexibility, such that model fitting becomes pathological due to increasing parameter degeneracies within the model. In the following sections, we will describe these layers, mapping functions, and occurrences of these fatal pathologies.

@@

For the moment, let us consider a single supernova, and represent its true parameters as $\theta$. The underlying supernova model and intrinsic dispersion model gives the probability of $p(y|\theta)$ for obtaining light curve data $y$. As mentioned earlier, we work with summary statistics -- numbers output from a light curve fitter given the input $y$. The function $f$ describes the action of the fitter $\lbrace \hat{m}_B, \hat{x}_1., \hat{c} \rbrace = f(y)$. For notational simplicity, let us denote $\eta \equiv \lbrace m_B, x_1, c \rbrace$. The model prediction for our derived data is

@@

Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift, and subtract this shift from our observed data. With the extra dispersion introduced from the $\kappa$ parameters addressing the under-reporting of error, shifting the mean colour address the non-Gaussianity introduced by chromatic smearing. We allow the amount of shift to be parametrised to account for different intrinsic scatter models where parameter $s_c$ controls the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations. The scatter models, sourced from \citet{Guy2010} and \citet{Chotard2011} are discussed in detail in Section \ref{sec:simdes}. The maximum shift was found to using the intrinsic scatter model from \citet{Chotard2011}. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a global shift in supernovae colour across all surveys. We place a uniform prior on $s_c$. It is important to note that the effects being modelled and characterised by simulations here are features of the supernova population, not cosmological features, and so the choice of fiducial cosmology has negligible impact.

@@

The underlying supernova population is characterised by underlying distributions in colour, stretch and absolute magnitude. Analytic models often treat the colour and stretch populations independently with a skew normal and normal, respectively. Intrinsic dispersion is treated in a variety of manners in other models, from representing it simply as (Gaussian) scatter on the absolute magnitude of the supernova population, to correlated multivariate normal scatter on the combined magnitude, stretch and colour distribution. Additionally, redshift drift of populations' mean colour and stretch will introduce a cosmological bias in our fits unless the model population possesses similar ability to change as a function of redshift.

We follow prior work and model the colour population as an independent redshift-dependent skew normal for each survey. For the stretch population, we adopt a redshift-dependent normal, and magnitude dispersion is modelled as a normal. Following {\rubin} we allow the mean colour and stretch to vary over redshift, anchoring four equally spaced redshift nodes spanning the redshift range of each survey, linearly interpolating between the nodes. Each set of nodes are anchored to equally spaced to span the minimum to maximum redshift of each survey. These nodes are represented as $\langle x_1^i \rangle$ and $\langle c^i \rangle$. Both the colour and stretch means are modelled with normal priors. Initial versions of our model adopted a fully covariant multivariate skew normal (with skewness set to zero only for the magnitude component), however pathological fitting complications required us to simplify our treatment. We parameterise the skewness $\alpha_c$ by sampling $\delta_c = \alpha_c / \sqrt{1 + \alpha_c^2}$ which itself is given a uniform prior $\mathcal{U}(0,0.98)$, which allows $\alpha_c$ to span positive values in the realm of physical plausibility as determined from constraints in \citet{Scolnic2016}. The width of the population, represented by the vector $\lbrace \sigma_{M_B}, \sigma_{x_1}, \sigma_c \rbrace$ is subject to Cauchy priors, however are sampled in log space for efficiency in sampling close to zero.

@@

On top of the SNIa populations, as described above, we also include a simplistic outlier population that also follows {\rubin} \citep[and therefore ][]{Kunz2007} as a Gaussian mixture; where the mean of the population is fixed to the SNIa population, but the population width is set to a width of $\sigma^{\rm outl} = 1$ in $M_B$, $x_1$ and $c$. With the spectroscopic DES sample, the contamination rate is expected to be far too low to actually fit contamination population, however in future works with photometric samples that will suffer from significantly more contamination, it will be required that extra degrees of freedom are afforded the outlier population. This contamination is not just a result of non-SNIa being observed, but can also arise from host galaxy misidentification and incorrect redshifts - depending on the cuts on the optimised ratio between purity and efficiency, this can result in between 3\% to 9\% misidentification  \citep{Gupta2016}. Proof of concept analysis on simple simulation data show that an acceptable parametrisation to deal with contamination, both from non type Ia supernovae and from catastrophic failure, is to represent the typically dimmer contaminant population as $\langle M_B^{\rm outl} \rangle = \langle M_B \rangle + \delta_{M_B}^{\rm outl}$, where $\delta_{M_B}^{\rm outl}$ is constrained to be positive, or even to be greater than a small positive number to reduce degeneracy between the two populations. For the purposes of the DES spectroscopic sample, which will be confirmed SNIa, $\delta_{M_B}^{\rm outl} =0$. We assume that supernovae fall into the SNIa population as determined by their observed classification probability $\hat{p}$, which is set to unity for the spectroscopic sample.

@@

It is now well known that host galaxy environment is significantly correlated with supernova properties. The latest sample of over 1300 spectroscopically confirmed Type Ia supernovae show $>5\sigma$ evidence for correlation between host mass and luminosity \citep{Uddin2017}. The traditional correction, as employed in analyses such as \citet{Suzuki2012} and \citet{Betoule2014} invoke a step function such that $\Delta M = 0.08 \mathcal{H}(\log(M) - 10))$, where $\mathcal{H}$ is the Heaviside step function, $M$ is the galaxy mass in solar masses and $0.08$ represents the size of the magnitude step. The scale of this step function varies from analysis to analysis, with the 0.08 value shown previously sourced from \cite{Sullivan2010} and used in \citet{Betoule2014}. In this work we adopt the model used in {\rubin}, which follows the work from \citet{Rigault2013}, such that we introduce two parameters to incorporate a redshift-dependent host galaxy mass correction:

\begin{align}
\Delta M = \delta(0) \left[ \frac{1.9\left(1 - \frac{\delta(0)}{\delta(\infty)}\right)  }{0.9 + 10^{0.95z}} + \frac{\delta(0)}{\delta(\infty)}\right], \label{eq:mass}
\end{align}

@@

This method of adjusting the observed summary statistics is used throughout the traditional and BHM analyses. For each survey and band, we have two systematics --  the calibration uncertainty and the filter wavelength uncertainty. We include these in our approach, in addition to including HST Calspec calibration uncertainty, ten SALT2 model systematic uncertainties, a dust systematic, a global redshift bias systematic, and also the systematic peculiar velocity uncertainty. This gives thirteen global systematics shared by all surveys, plus two systematics per band in each survey. With $\eta \equiv \lbrace m_B, x_1, c \rbrace$, our initial conditional likelihood for our observed summary statistics shown in Equation \eqref{eq:pop} becomes
